To be FAIR: Theory Specification Needs an Update

### 2025-02-21 manuscript.Rmd
update 2025-02-21 11:33:31.377054

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
FAIR theory has the potential to improve the efficiency of scholarly communication and
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences, and beyond.

@@

The present paper introduces open science methods for theory specification and archival.
Merely publishing theory in a research article does not make it open;
to be open, theory should adhere to established open science standards.
We do so by implementing theories as information artifacts,
archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
FAIR theories are *Findable* via a DOI or search,
*Accessible* in a machine- and human-readable filetype,
*Interoperable* within the data analysis environment,
and *Reusable* in the practical and legal sense, so that they may be iteratively improved - potentially via a participative process.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see Table \@ref(tab:tabfair).
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

The four Findability criteria are applicable to theory with only minor adjustments, see Table \@ref(tab:tabfair).
First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
Of the many services that provide DOIs for scientific information artefacts,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata	(F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs [@boscoMetaBUSVehicleFacilitating2017],
ontology [@guyonMeasurementOntologyEpistemology2018],
or catalog of psychological phenomena [REF Noah Denny].
Metadata should also include identifiers for all the versions of the theory it describes	(F3);
Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.

Finally, metadata should be registered or indexed in a searchable registry (F4).
It is important to note that, while may archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
only few are specifically designed for FAIR-compliant archival.
Zenodo stands out in this respect.
Thus, while it is fine (and recommended) to use GitHub for version control,
the version of record should be archived in a FAIR repository like Zenodo.
Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
Furthermore, we suggest using the keyword `"fairtheory"` for all resources that constitute or reference (a) FAIR theory.
<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->

Findability is substantially amplified if intended users of a resource know where to search for it.
This is a known problem in relation to research data and software [@katzSpecialIssueSoftware2024].
Regrettably, most academic search engines are designed to index traditional print publications, not other information artifacts.
Since the status quo is to publish theories in papers,
the FAIR requirements are met if scholars continue to do so,
and additionally publish theories as separate information artifacts.
The `"fairtheory"` keyword can also be used to signal the presence of theory within a paper.
<!-- An ad-hoc solution is to publish a paper as documentation for the information artifact [@mcgillivrayDeepImpactStudy2022]. -->
<!-- There is even a journal - the [Journal of Open Source Software](https://joss.theoj.org/) - that publishes software documentation as a paper so that it will be indexed by search engines. -->
<!-- The same solution could be applied to theories. -->
In the longer term, it may not be necessary to write a paper for each theory.
If Zenodo becomes more recognized as centralized repository for information artifacts, researchers may begin to search there more regularly.
Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index information artifacts stored in Zenodo.

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [@grayHowMapTheory2017].
Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability, however.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.

@@

Interoperability pertains to the property of information artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
Firstly, theory and its associated metadata should use a formal, Accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable datatype,
as previously recommended [@vanlissaWORCSWorkflowOpen2021].
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some formal language.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps [@grayHowMapTheory2017].
While a theory represented as a bitmap image is not very Interoperable,
the same image represented in the DOT language [@DOTLanguage]<!--AB citation needed; also Aaron has his own Julia package for graph language specs - not sure whether Aaron wants to refer to this?--> for representing graphs does meet this ideal.

@@

If we take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
Applied to FAIR theory, reusability requires that  each theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and Accessible license for reuse (R1.1).
It should further have detailed provenance (R1.2), 
which is achieved through version control with Git and archival on Zenodo.
Finally, the (meta)data which meets domain-relevant community standards (R1.3).
The Datacite metadata schema offers an initial template in this regard,
and this paper takes one step towards establishing more fine-grained community standards for FAIR theory.

@@

The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
In particular, the practice of version control offers extensive benefits for scientific work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
To version control our project, we initiate a Git repository in the project folder.
<!-- This can be done by [installing Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) locally, -->
<!-- opening a terminal in the project folder, -->
<!-- and running the command `git init`. -->
We subsequently create a remote repository to host a copy of this local Git repository on GitHub, which will in turn be archived.
<!-- To [do this on GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/), create an account and press the button labeled `Create new repository`. -->
<!-- The default settings for the repository are fine; -->
<!-- for the next steps of the tutorial it  -->
Note that the repository
must be set to "Public" to take advantage of GitHub's Zenodo integration.
<!-- Copy the repository's URL, which usually has the form `https://github.com/username/repository.git`. -->
<!-- Next, in the terminal window, run the following lines of code sequentially: -->

@@

Note that this function also automatically provides basic FAIR theory metadata to Zenodo.

>>>>

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
FAIR theory has the potential to improve the efficiency of scholarly communication and
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences, and beyond.

@@

The present paper introduces open science methods for theory specification and archival.
Merely publishing theory in a research article does not make it open;
to be open, theory should adhere to established open science standards.
We do so by implementing theories as information artifacts,
archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are "data about the data", they provide information about the nature and content of an information artifact.
Metadata are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
*Accessible* in a machine- and human-readable filetype;
*Interoperable* for specific purposed, for example, within the data analysis environment;
and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see Table \@ref(tab:tabfair).
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

The four Findability criteria are applicable to theory with only minor adjustments, see Table \@ref(tab:tabfair).
First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
Of the many services that provide DOIs for scientific information artifacts,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata	(F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs [@boscoMetaBUSVehicleFacilitating2017],
ontology [@guyonMeasurementOntologyEpistemology2018],
or catalog of psychological phenomena [REF Noah Denny].
Metadata should also include identifiers for all the versions of the theory it describes	(F3);
Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.

Finally, metadata should be registered or indexed in a searchable registry (F4).
It is important to note that, while may archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
only few are specifically designed for FAIR-compliant archival.
Zenodo stands out in this respect.
Thus, while using Git for version control and GitHub for collaboration has specific advantages for scientific work [@ramGitCanFacilitate2013],
the version of record should be archived in a FAIR repository like Zenodo.
Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
Furthermore, we suggest using the keyword `"fairtheory"` for all resources that constitute or reference (a) FAIR theory.
<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->

Findability is substantially amplified if intended users of a resource know where to search for it.
This is a known problem in relation to research data and software [@katzSpecialIssueSoftware2024].
Regrettably, most academic search engines only index traditional print publications, not other information artifacts.
Since the status quo is to publish theories in papers,
the FAIR requirements are met if scholars continue to do so,
and additionally publish theories as separate information artifacts.
The `"fairtheory"` keyword can also be used to signal the presence of theory within a paper.
<!-- An ad-hoc solution is to publish a paper as documentation for the information artifact [@mcgillivrayDeepImpactStudy2022]. -->
<!-- There is even a journal - the [Journal of Open Source Software](https://joss.theoj.org/) - that publishes software documentation as a paper so that it will be indexed by search engines. -->
<!-- The same solution could be applied to theories. -->
In the longer term, it may not be necessary to write a paper for each theory.
If Zenodo becomes more recognized as centralized repository for information artifacts, researchers may begin to search there more regularly.
Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index information artifacts stored in Zenodo.

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [@grayHowMapTheory2017].
Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
for example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places and prevent inconsistencies from arising.

@@

Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
Firstly, theory and its associated metadata should use a formal, Accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable datatype,
as previously recommended [@vanlissaWORCSWorkflowOpen2021].
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some formal language.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps [@grayHowMapTheory2017].
While a theory represented as a bitmap image is not very Interoperable,
the same image represented in the DOT language [@DOTLanguage]<!--AB citation needed; also Aaron has his own Julia package for graph language specs - not sure whether Aaron wants to refer to this?--> for representing graphs does meet this ideal.

@@

If we take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
Applied to FAIR theory, reusability requires that  each theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and Accessible license for reuse (R1.1).
It should further have detailed provenance (R1.2), 
which is achieved through version control with Git and archival on Zenodo.
Finally, the (meta)data which meets domain-relevant community standards (R1.3).
The Datacite metadata schema offers an initial template in this regard,
and this paper takes one step towards establishing more fine-grained community standards for FAIR theory.
[This is an example of FAIR metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/DGPSabstract.md) extracted from Zenodo.

@@

The field of computer science provides well-established processes for creating information artifacts that can be iteratively improved.
In particular, the practice of version control offers extensive benefits for scientific work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
To version control our project, we initiate a Git repository in the project folder.
<!-- This can be done by [installing Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) locally, -->
<!-- opening a terminal in the project folder, -->
<!-- and running the command `git init`. -->
We subsequently create a remote repository to host a copy of this local Git repository on GitHub, which will in turn be archived.
<!-- To [do this on GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/), create an account and press the button labeled `Create new repository`. -->
<!-- The default settings for the repository are fine; -->
<!-- for the next steps of the tutorial it  -->
Note that the repository
must be set to "Public" to take advantage of GitHub's Zenodo integration.
<!-- Copy the repository's URL, which usually has the form `https://github.com/username/repository.git`. -->
<!-- Next, in the terminal window, run the following lines of code sequentially: -->

@@

Note that this function also automatically submits basic FAIR theory metadata to Zenodo.
